Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-08 _tex/index.tex
Built site for gh-pages

@@

In statistics, one application incorporating concepts from graph theory
is causal inference. Specifically, the structural approach to causal
inference uses directed acyclic graphs (DAG) to provide a graphical and
formal representation of the causal structure of a system
\citep{Neal_2020}. In this context, a \emph{graph} denotes a collection
of nodes connected by edges, where nodes represent random variables. The
term \emph{directed} indicates the edges of the graph extend from one
node to another, with arrows showing the direction of causal influence.
Moreover, the term \emph{acyclic} indicates the causal influences do not
form a loop, meaning the influences do not cycle back on themselves
\citep{McElreath_2020}. Several examples of DAGs can be found in
Figure~\ref{fig-dags_scms}.

@@

Often, DAGs omit the exogenous variables for simplicity, resulting in
the \emph{standard} representation. However, including exogenous
variables in a graph can be beneficial in some scenarios, as their
presence can reveal potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}, concepts explored in the
following section. The standard representation is illustrated in the
middle panels of Figure~\ref{fig-dags_scms}.

Lastly, the right panels of Figure~\ref{fig-dags_scms} depict the SCM
form of the fundamental building blocks. SCMs are formal mathematical
models defined by a set of endogenous variables \(V\), a set of
exogenous variables \(E\), and a set of functions
\(F=\{f_{X},f_{Z},f_{Y}\}\) \citep{Pearl_2009, Neal_2020}. These
functions, referred to as structural equations, specify each endogenous
variable as nonparametric functions of other variables. Moreover, SCMs
use the symbol `\(:=\)' to indicate the variables' asymmetrical causal
dependence and the symbol `\(\perp\!\!\!\perp\)' to represent
\emph{d-separation}, which roughly equates to the concept of variable
independence. The concepts of d-separation and causal (in)dependence are
explored in the following section.

@@

Given the heuristic nature of DAGs, a motivating example can help
clarify how to use the five fundamental building blocks to construct the
causal structure of a system. Consider a research problem where the
causal effect of a variable \(T\) on an outcome \(Y\) needs
investigation. The problem suggests that a variable \(X\) potentially
influences both \(T\) and \(Y\). Beyond these relationships, the problem
does not specify any further variables of interest. Such scenarios are
not hard to imagine. For instance, \(T\) might represent different
treatments that could affect the recovery from cancer \(Y\), while \(X\)
could denote the patient's age. Similarly, in the context of a DCJ study
like the one described by \citet{Boonen_et_al_2020}, \(T\) could
represent the duration of a child's cochlear implant use, which might
influence the child's overall speech quality \(Y\), with \(X\)
indicating the child's hearing status. {(not a bad example, but I prefer
one using writing skills)}

Figure~\ref{fig-example} illustrates the plausible causal structure of
the motivating example. A detailed examination of Figures
\ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM
\ref{fig-scm_example1} reveals the presence of at least four of the five
fundamental building blocks. The figures display multiple descendants,
as indicated by pairwise relations such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). Additionally, the figures
features multiple pairs of unconnected nodes, evident from the relations
\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate the
fork \(X \rightarrow \{T,Y\}\), and two colliders with
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).

\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_example1.png}

\subcaption{\label{fig-mdag_example1}Magnified representation}

\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_example1.png}

\subcaption{\label{fig-sdag_example1}Standard representation}

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := f_{Z}(X,e_{T}) \\
  Y & := f_{Y}(T,X,e_{Y}) \\
  e_{T} & \perp\!\!\!\perp e_{X} \\
  e_{T} & \perp\!\!\!\perp e_{Y} \\
  e_{X} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

\subcaption{\label{fig-scm_example1}Structural causal model}

\caption{\label{fig-example}DAGs for a plausible causal structure in a
system.}

>>>>

\subsection{The empirical research
endeavor}\label{sec-framework-empirical}

The goal of an investigator is to generate new knowledge or deepen the
understanding of specific topics or issues. Investigators accomplish
this by addressing research problems. Research problems fall into two
categories: empirical and non-empirical research. The first tackles
real-world challenges, relying on evidence gathered through observation
and experimentation. The latter focuses on conceptual inquiries that
enhance understanding complex concepts, theories, and phenomena.

Addressing research problems requires formulating well-defined research
questions that provide focus, clarity, and structure to a study. In
empirical research, these questions are often framed as \emph{targets of
inference} or \emph{estimands}, representing the quantities the
investigator aims to estimate. For instance, an investigator might want
to examine how the duration of cochlear implant use affects the average
overall speech quality of a child.

Furthermore, in empirical research, investigators must design
appropriate methodologies, collect data, and ensure a systematic and
coherent analysis of the results. To achieve this, they must accurately
identify the factors influencing an outcome and correctly attribute the
effects to each factor. Consequently, understanding how an investigator
transitions from a target of inference to generating statistical
estimates is crucial.

Figure~\ref{fig-IEflow} shows the Identification-Estimation flowchart
\citep{McElreath_2020, Neal_2020}, which illustrates the mechanism for
transitioning from a estimand to a corresponding \emph{statistical
estimate}, through identification and estimation.

@@

In statistics, one application incorporating concepts from graph theory
is causal inference. Specifically, the structural approach to causal
inference uses directed acyclic graphs (DAG) to provide a graphical and
formal representation of the causal structure of a system
\citep{Neal_2020}. In this context, a \emph{graph} denotes a collection
of nodes connected by edges, where nodes represent random variables. The
term \emph{directed} indicates the edges of the graph extend from one
node to another, with arrows showing the direction of causal influence.
Moreover, the term \emph{acyclic} indicates the causal influences do not
form a loop, meaning the influences do not cycle back on themselves
\citep{McElreath_2020}.

@@

Often, DAGs omit the exogenous variables for simplicity, resulting in
the \emph{standard} representation. However, including exogenous
variables in a graph can be beneficial in some scenarios, as their
presence can reveal potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}, concepts explored in
Section~\ref{sec-framework-flow}. The standard representation is
illustrated in the middle panels of Figure~\ref{fig-dags_scms}.

Lastly, the right panels of Figure~\ref{fig-dags_scms} depict the SCM
form of the fundamental building blocks. SCMs are formal mathematical
models defined by a set of endogenous variables \(V\), a set of
exogenous variables \(E\), and a set of functions
\(F=\{f_{X},f_{Z},f_{Y}\}\) \citep{Pearl_2009, Neal_2020}. These
functions, referred to as structural equations, specify each endogenous
variable as nonparametric functions of other variables. Moreover, SCMs
use the symbol `\(:=\)' to indicate the variables' asymmetrical causal
dependence and the symbol `\(\perp\!\!\!\perp\)' to represent
\emph{d-separation}, which roughly equates to the concept of variable
independence. The concepts of d-separation and causal (in)dependence are
explored in Section~\ref{sec-framework-flow}.

@@
